Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-06-18 vertical-thermal-review.Rmd
addressing comment about gs

Leaf hydraulic, stomatal and boundary layer conductances are all critical for regulating *T~leaf~* via latent heat loss (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), *g~b~* increases with height (Table `r table_leaf_metabolism`).
Maximum *g~s~* increases with light -- typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979] -- and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands caused by the high radiation and *g~b~* in sun leaves with fully open stomata, in part because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014]. 
These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ (section 3.3, Table `r table_leaf_traits`).
Stomatal depression reduces transpirational cooling (Fig. `r fig_leaf_T`e), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which *g~s~* is greatest, *T~opt~* of *g~s~*, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in *g~s~* in response to rising *T~air~* (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. 
This, added to the tendency for sun leaves to have higher *T~leaf~*, implies that high *T~air~* should decrease *g~s~* of upper canopy leaves more than understory leaves, particularly when water availability is limited.

>>>>

Leaf hydraulic, stomatal and boundary layer conductances are all critical for regulating *T~leaf~* via latent heat loss (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), *g~b~* increases with height (Table `r table_leaf_metabolism`).
Maximum *g~s~* increases with light -- typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979] -- and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands caused by the high radiation and *g~b~* in sun leaves with fully open stomata, in part because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014]. 
These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ (section 3.3, Table `r table_leaf_traits`).
Stomatal depression reduces transpirational cooling (Fig. `r fig_leaf_T`e), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which *g~s~* is greatest, *T~opt~* of *g~s~*, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in *g~s~* in response to rising *T~air~* (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. 
Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per *se* in nature, causing *g~s~* to decline when the leaf warms.
